Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-07 pnas/Towards.tex
additional editings

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_1}
		\caption{$500$ micrographs \TODO{Tamir: I am redoing the exp}}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_4}
		\caption{$2\times 10^3$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_40}
		\caption{$2\times 10^4$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_400}
		\caption{$2\times 10^5$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $500,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels), each containing $700$ image occurrences on average.}	
\end{figure*}

@@

In general, for a signal $z$ of length $m$, the autocorrelation of order $q = 1, 2, \ldots$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{align}
a_z^q[\ell_1,\ldots,\ell_{q-1}]  & = \frac{1}{m} \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}],
\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the range $0, \ldots, m-1$ is zero-padded.
For our purposes, this will be applied both to $x$ (of length $L$) and to $y$ (of length $N$).
Explicitly, the first-, second- and third-order autocorrelations are given by: %\TODO{we may want to remove these explicit formulas}
\begin{align} 
a_z^1 & = \frac{1}{m} \sum_{i=0}^{m-1} z[i], \nonumber\\
a_z^2[\ell] & = \frac{1}{m} \sum_{i = \max\{0, -\ell\}}^{m-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
a_z^3[\ell_1,\ell_2] & = \frac{1}{m} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{m-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
The autocorrelation functions have symmetries. Specifically, $a_z^2[\ell] = a_z^2[-\ell]$, and
$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
$

@@

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph.
%
\begin{proposition} \label{prop:gamma}
	Let $\sigma > 0$ be fixed and assume that the separation condition~\eqref{eq:spacing} holds. If the mean of $x$ is nonzero, then 
	%
	\begin{align*}
	%
	\gamma & \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	%
	\end{align*}
	%
\end{proposition}
\begin{proof}
In the limit, $(a^1_y)^2=\frac{\gamma^2}{L^2}\sum_{i=0}^{L-1}\sum_{j=0}^{L-1}x[i]x[j]$. Similarly,  $\sum_{\ell = 1}^{L-1}a_y^2[\ell]=\frac{\gamma}{L}\sum_{\ell = 1}^{L-1}\sum_{i=0}^{L-1-\ell}x[i]x[i+\ell]$ and $a_y^2[0]=\frac{\gamma}{L}\sum_{i=0}^{L-1}x^2[i] + \sigma^2$. The proof is concluded by noting that  $a_x^2[-\ell]=a_x^2[\ell]$. 
%	Plug in the explicit expressions of~\eqref{eq:mean_micrograph} and~\eqref{eq:ac2_micrograph} into the right hand side of the equality.
\end{proof}

Using third-order autocorrelation information of $y$, both the ratio $\gamma$ and the noise $\sigma$ are determined. For the following results, when we say that a result holds for a ``generic'' signal $x$, we mean that the set of signals which cannot be determined by these measurements
has Lebesgue measure zero. 
In particular, this means that we can recover
almost all signals with the given measurements.

\begin{proposition} \label{prop:gamma_sigma}
	%
	%Let $\sigma > 0$ be fixed. 
	Assume $L \geq 3$ and assume that the separation condition~\eqref{eq:spacing} holds. % Added by Nicolas, Sep 4, 2018: I checked all identities for L >= 3, with some computer code; I didn't write code to verify the corner cases L = 1 and L = 2.
	In the limit of $N\to \infty$, the observed autocorrelations $a_y^1,a_y^2$ and  $a_y^3$ determine the ratio $\gamma$ and noise level $\sigma$ uniquely for a generic signal $x$. If $\gamma > \frac{1}{4}$, then this holds for any signal $x$ with nonzero mean. 
	\begin{proof}
		See Appendix~\ref{sec:proof_prop_gamma_sigma}.
	\end{proof}
\end{proposition}

From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	 In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if:
	\begin{itemize}
		\item Either the signal $x$ is generic; or
		\item   $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
	\end{itemize}
\end{corollary}
%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

The technique we advocate allows recovery of a signal hidden in noisy micrographs without detecting the location  of the signals embedded in these micrographs. To illustrate the underlying principles of the method, we present several numerical examples for the toy model~\eqref{eq:model} and a simple proof of concept for simulated cryo-EM data.  Appendix~\ref{sec:numeric_details} provides additional details on the experiments. 
\revised{The code to generate all figures is publicly available in TKTK.}

@@

>>>>

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_1}
		\caption{$500$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_4}
		\caption{$2\times 10^3$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_40}
		\caption{$2\times 10^4$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_400}
		\caption{$2\times 10^5$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $500,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels), each containing $700$ image occurrences on average.  \TODO{Tamir: I am redoing the exp}}	
\end{figure*}

@@

In general, for a signal $z$ of length $m$, the autocorrelation of order $q = 1, 2, \ldots$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
\begin{align}
a_z^q[\ell_1,\ldots,\ell_{q-1}]  & = \frac{1}{m} \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}],
\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the range $0, \ldots, m-1$ is zero-padded.
For our purposes, this will be applied both to $x$ (of length $L$) and to $y$ (of length $N$).
%This part can be added to an arxiv version of the manuscript
%Explicitly, the first-, second- and third-order autocorrelations are given by: %\TODO{we may want to remove these explicit formulas}
%\begin{align} 
%a_z^1 & = \frac{1}{m} \sum_{i=0}^{m-1} z[i], \nonumber\\
%a_z^2[\ell] & = \frac{1}{m} \sum_{i = \max\{0, -\ell\}}^{m-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
%a_z^3[\ell_1,\ell_2] & = \frac{1}{m} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{m-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
%\end{align}
%The autocorrelation functions have symmetries. Specifically, $a_z^2[\ell] = a_z^2[-\ell]$, and
%$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
%$

@@

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph. The proof is provided in Appendix~\ref{sec:proof_prop_gamma}.
%
\begin{proposition} \label{prop:gamma}
	Let $\sigma > 0$ be fixed and assume that the separation condition~\eqref{eq:spacing} holds. If the mean of $x$ is nonzero, then 
	%
	\begin{align*}
	%
	\gamma & \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	%
	\end{align*}
	%
\end{proposition}
%\begin{proof}
%In the limit, $(a^1_y)^2=\frac{\gamma^2}{L^2}\sum_{i=0}^{L-1}\sum_{j=0}^{L-1}x[i]x[j]$. Similarly,  $\sum_{\ell = 1}^{L-1}a_y^2[\ell]=\frac{\gamma}{L}\sum_{\ell = 1}^{L-1}\sum_{i=0}^{L-1-\ell}x[i]x[i+\ell]$ and $a_y^2[0]=\frac{\gamma}{L}\sum_{i=0}^{L-1}x^2[i] + \sigma^2$. The proof is concluded by noting that  $a_x^2[-\ell]=a_x^2[\ell]$. 
%%	Plug in the explicit expressions of~\eqref{eq:mean_micrograph} and~\eqref{eq:ac2_micrograph} into the right hand side of the equality.
%\end{proof}

Using third-order autocorrelation information of $y$, both the ratio $\gamma$ and the noise $\sigma$ are determined. For the following results, when we say that a result holds for a ``generic'' signal $x$, we mean that the set of signals which cannot be determined by these measurements
has Lebesgue measure zero. 
In particular, this means that we can recover
almost all signals with the given measurements. The proof is provided in Appendix~\ref{sec:proof_prop_gamma_sigma}.

\begin{proposition} \label{prop:gamma_sigma}
	%
	%Let $\sigma > 0$ be fixed. 
	Assume $L \geq 3$ and assume that the separation condition~\eqref{eq:spacing} holds. % Added by Nicolas, Sep 4, 2018: I checked all identities for L >= 3, with some computer code; I didn't write code to verify the corner cases L = 1 and L = 2.
	In the limit of $N\to \infty$, the observed autocorrelations $a_y^1,a_y^2$ and  $a_y^3$ determine the ratio $\gamma$ and noise level $\sigma$ uniquely for a generic signal $x$. If $\gamma > \frac{1}{4}$, then this holds for any signal $x$ with nonzero mean. 
%	\begin{proof}
%		See Appendix~\ref{sec:proof_prop_gamma_sigma}.
%	\end{proof}
\end{proposition}

From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
\begin{corollary}
	%Let $K=1$ and $\sigma > 0$ be fixed.
	In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if either the signal $x$ is generic or $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
\end{corollary}

%This can be added to an arxiv version of the paper	
%From Propositions~\ref{prop:uniqueness} and~\ref{prop:gamma_sigma} we   deduce the following:
%\begin{corollary}
%	%Let $K=1$ and $\sigma > 0$ be fixed.
%	 In the limit of $N\to \infty$ and under the separation condition~\eqref{eq:spacing}, the signal $x$, the ratio $\gamma$, and the noise level $\sigma$ are determined from the first three autocorrelation functions of $y$ if:
%	\begin{itemize}
%		\item Either the signal $x$ is generic; or
%		\item   $x$ has nonzero mean  and $\gamma > \frac{1}{4}$.
%	\end{itemize}
%\end{corollary}
%
%\TODO{Similar results hold for signals in higher dimensions.---need to keep this sentence in red until the very end to make sure it's correct.}

@@

The technique we advocate allows recovery of a signal hidden in noisy micrographs without detecting the location  of the signals embedded in these micrographs. To illustrate the underlying principles of the method, we present several numerical examples for the toy model~\eqref{eq:model} and a simple proof of concept for simulated cryo-EM data.  Appendix~\ref{sec:numeric_details} provides additional details on the experiments. 
The code to generate all figures is publicly available in \url{https://github.com/PrincetonUniversity/BreakingDetectionLimit}.

@@

\section{Proof of Proposition~\ref{prop:gamma}} \label{sec:proof_prop_gamma}

In the limit, 
\begin{equation*}
(a^1_y)^2=\frac{\gamma^2}{L^2}\sum_{i=0}^{L-1}\sum_{j=0}^{L-1}x[i]x[j].
\end{equation*}
 Similarly,  
 \begin{equation*}
\sum_{\ell = 1}^{L-1}a_y^2[\ell]=\frac{\gamma}{L}\sum_{\ell = 1}^{L-1}\sum_{i=0}^{L-1-\ell}x[i]x[i+\ell],
 \end{equation*}
and $a_y^2[0]=\frac{\gamma}{L}\sum_{i=0}^{L-1}x^2[i] + \sigma^2$. The proof is concluded by noting that  $a_x^2[-\ell]=a_x^2[\ell]$. 
%%	Plug in the explicit expressions of~\eqref{eq:mean_micrograph} and~\eqref{eq:ac2_micrograph} into the right hand side of the equality.
